---
title: "HW_1"
output: html_document
date: "2024-06-24"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Libraries and setting working directory


```{r Chunk_1}
library(tidyverse)
library(ggplot2)
folder_path <- "/Users/ivanzamorano/Desktop/HW1/Datasets" # use your own path here
setwd(folder_path) # passing the object into the function
```

## Reading in the files


```{r Chunk_2}
people <- read_csv("Traffic_Crashes_-_People.csv")
crashes <- read_csv("Traffic_Crashes_-_Crashes.csv")
fert <- read_csv("wave_3_fertility.csv")
power <- read_csv("wave_3_power.csv")

```

## Question 1 ~ 2

Among female sex individuals of age 65 and above in this dataset, what proportion are pedestrians?

```{r Chunk_3}

people_filtered1 <- people %>%
  filter(SEX == 'F') %>%
  filter(AGE >= 65)

person_type1 <- table(people_filtered1$PERSON_TYPE)
# Convert the frequency table into proportions

person_type_proportions <- prop.table(person_type1)

# Extract the proportion of pedestrians

pedestrian_proportion <- person_type_proportions["PEDESTRIAN"]

pedestrian_proportion

```

## Question  3 ~ 4

Filter the dataset to only those with no indication of injury. How old is the oldest person in this filtered dataset? Use the INJURY_CLASSIFICATION and AGE columns to answer this question.


```{r Chunk_4}
people_filtered2 <- people %>%
  filter(INJURY_CLASSIFICATION == 'NO INDICATION OF INJURY')

max_age <- max(people_filtered2$AGE, na.rm = TRUE)
max_age
```



## Question 5 ~ 6
Again using the dataset entitled “Traﬀic_Crashes_-_People.csv”, answer the following.
How many people in the dataset are age 25 or younger and are passengers? Use PERSON_TYPE and AGE to answer this question.

```{r Chunk_5}
people_filtered3 <- people %>%
    filter(AGE <= 25 & PERSON_TYPE == "PASSENGER")

count_25_younger_passenger<- nrow(people_filtered3)
count_25_younger_passenger
```


## Question 7 ~ 8

Begin with the “Traﬀic_Crashes_-_People.csv” dataset.
• Filter for a subset of observations pertaining to individuals who are age 65 or older. • Next, Join with the crashes dataset.
• Get the year from CRASH_DATE.
How many distinct people age 65 or above were involved in a car crash during the year 2020?


```{r Chunk_6}

people_filtered4 <- people %>%
    filter(AGE >= 65)

merged_df <- merge(people_filtered4, crashes, by = "CRASH_RECORD_ID")

# Parse the CRASH_DATE to ensure it is in date-time format
merged_df$CRASH_DATE <- as.POSIXct(merged_df$CRASH_DATE, format = "%m/%d/%Y %I:%M:%S %p")

# Create a new variable CRASH_DATE_YEAR that contains only the year
merged_df$CRASH_DATE_YEAR <- format(merged_df$CRASH_DATE, "%Y")

merged_df_2020 <- merged_df %>%
  filter(CRASH_DATE_YEAR == "2020")

distinct_people_count <- merged_df_2020 %>%
  summarise(count = n_distinct(PERSON_ID)) %>%
  pull(count)

distinct_people_count
```
## Question 9

Your task is to plot the number of people who died in car crashes over time (every year) and submit an image of your final result in a lineplot, as well as your code in the following question.

• Begin with the people dataset, and keep only the observations where the variable IN- JURY_CLASSIFICATION indicates a fatal injury.
• Next, join with the crashes dataset and get the year associated with each person who got into fatal car accident.
• Use ggplot to make a lineplot which illustrates the number of people who died in car crashes every year.

```{r Chunk_7}
merged_df <- merge(people, crashes, by = "CRASH_RECORD_ID") %>%
    filter(INJURY_CLASSIFICATION == "FATAL")

# Parse the CRASH_DATE to ensure it is in date-time format
merged_df$CRASH_DATE <- as.POSIXct(merged_df$CRASH_DATE, format = "%m/%d/%Y %I:%M:%S %p")

# Create a new variable CRASH_DATE_YEAR that contains only the year
merged_df$CRASH_DATE_YEAR <- format(merged_df$CRASH_DATE, "%Y")

death_counts <- table(merged_df$CRASH_DATE_YEAR)

# Convert to data frame for plotting
death_counts_df <- as.data.frame(death_counts)
death_counts_df$Year <- as.integer(rownames(death_counts_df))

# Plot the number of deaths over time in a line plot
ggplot(data = death_counts_df, aes(x = Year, y = Freq)) +
  geom_line() +
  geom_point() +
  labs(x = "Year", y = "Number of Deaths", title = "Number of Deaths in Car Crashes Over Time") +
  theme_minimal()

```


## Question 10 ~ 11

Note: For this question, please follow the instructions in the HW description to download and load the dataset: “wave_3_fertility.csv”. Spend some time exploring the names of the different columns of the datasets, and how they are organized. For many of the questions below, make sure to refer to the codebook linked in page 1.

Among women age 40 and above who have given birth to a child in this dataset, what proportion have never given birth to a girl?

Hint: Look at columns entitled age, evergivenbirth, and borngirls. After selecting in the dataset only observations who have given birth and age equal to or above 40, calculate what for what proportion borngirls is zero.

```{r Chunk_8}

fert_filtered <- fert %>%
  filter(evergivenbirth == 1) %>%
  filter(age >= 40)

count_borngirls_zero <- sum(fert_filtered$borngirls == 0, na.rm = TRUE)
count_borngirls_NA <- sum(is.na(fert_filtered$borngirls))

count_borngirls = count_borngirls_zero + count_borngirls_NA

count_borngirls

```

## Question 12 ~ 13

Again using the dataset entitled “wave_3_fertility.csv”, answer the following.
How many women in the dataset are age 20 or younger and have been pregnant in the past 12 months? Use pregnantlastyear and age to answer this question.

```{r Chunk_9}

fert_filtered <- fert %>%
  filter(age <= 20)

pregnant_12_months_zero <- sum(fert_filtered$pregnantlastyear == 0, na.rm = TRUE)
pregnant_12_months_NA<- sum(is.na(fert_filtered$pregnantlastyear))

pregnant_12_months = pregnant_12_months_NA + pregnant_12_months_zero

pregnant_12_months

```

## Question 14 ~ 15

Begin with the “wave_3_fertility.csv” dataset.

• Filter to observations of women that have children (borntotal >= 0).
• Next, join with the “wave_3_power.csv” dataset along two columns, FPrimary and
hhmid. These correspond to household and household member respectively.
• Count the number of rows
How many respondents with only daughter(s) believe it is better to send a son to school than a daughter (bettersontoschool = 1)? What if the respondent has only son(s)?

```{r Chunk_10}

# Filter to observations of women that have children (borntotal >= 0).

fert_filtered <- fert %>%
  filter(borntotal >= 0)

# Join with the “wave_3_power.csv” dataset along two columns, FPrimary and hhmid

merged_df <- merge(fert_filtered, power, by = c("FPrimary", "hhmid"), all = TRUE)

#Count the number of row

num_rows <- nrow(merged_df)
num_rows

merged_df_daughter <- merged_df %>%
  filter(bornboys == 0)

bettersontoschool_daughters <- sum(merged_df_daughter$bettersontoschool == 1, na.rm = TRUE)

bettersontoschool_daughters

fert_filtered <- fert %>%
  filter(borntotal >= 0)
merged_df <- merge(fert_filtered, power, by = c("FPrimary", "hhmid"), all = TRUE)

merged_df_daughter <- merged_df %>%
  filter(bornboys == 0)

merged_df_boys <- merged_df %>%
  filter(!is.na(bornboys) & bornboys != 0)

bettersontoschool_boys <- sum(merged_df_boys$bettersontoschool == 1, na.rm = TRUE)

bettersontoschool_boys

```

## Question 16

Your task is to plot a histogram of the number of children born to women who can no longer have children and submit an image of your final result, as well as your code in the following question.
• Begin with the fertility dataset and keep only the observations where the variable agemenopause is not equal to -1.
• Use ggplot to make a scatterplot which shows number of children born to women who are different ages. You must properly label your graph.
• You must include a title and labels on the x and y axes and pick a fill color of your choice for the bars.


```{r Chunk_11}

# Filter the observations where the variable agemenpause is not equal to -1

fert_filtered <- fert %>%
  filter(agemenopause != -1)


# Create scatterplot
ggplot(fert_filtered, aes(x = age, y = borntotal)) +
  geom_point(size = 3, color = "blue") +
  labs(x = "Woman's Age",
       y = "Number of Children Born",
       title = "Relationship Between Woman's Age and Number of Children") +
  theme_minimal()



# Create histogram
ggplot(fert_filtered, aes(x = borntotal)) +
  geom_histogram(binwidth = 1, fill = "skyblue", color = "black") +
  labs(x = "Number of Children Born",
       y = "Frequency",
       title = "Distribution of Number of Children Born to Women by Age") +
  facet_wrap(~age, scales = "free") +
  theme_minimal()


```
